---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r echo=FALSE}
library(dplyr)
```

## Loading and preprocessing the data

The first step is to load and preprocess the data.  Unzip the file to begin:

```{r echo=TRUE}
unzip("activity.zip")
```

Then, load the data into a data frame call `activity`:

```{r echo=TRUE}
activity <- read.csv("activity.csv")
```

Check the data frame to see if we need any processing or transformation:

```{r echo=TRUE}
str(activity)
```

It looks like the date column is stored as a factor, not a date.  Let's convert that to a date and group the data by date in a new data frame.  This used the `dplyr` package that is loaded without displaying to the screen.

```{r echo=TRUE}
activity$date <- as.Date(activity$date)

activity_by_date <- group_by(activity, date)
```

## What is mean total number of steps taken per day?

First, let's create a variable, `total_steps`,  holding the total number of steps taken on each day.  We also rename the column header on the second column.

```{r echo=TRUE}
total_steps <- summarize(activity_by_date, sum(steps, na.rm = TRUE))
names(total_steps)[2] <- "sum"

total_steps
```

Here is a histogram of the total steps each day:

```{r echo=TRUE}
hist(total_steps$sum, xlab = "Total Steps on Each Day", main = "Histogram of Total Steps on Each Day")
```

The mean number of steps each day is:

```{r echo=TRUE}
mean(total_steps$sum)
```

The median number of steps each day is:

```{r echo=TRUE}
median(total_steps$sum)
```

## What is the average daily activity pattern?

Let's return to the original data frame and group on 5-minute interval this time.  Again, this uses the `dplyr` package.

```{r echo=TRUE}
activity_by_interval <- group_by(activity, interval)
```

Now, put the mean for each interval into a new variable, `mean_steps` and rename the mean column:

```{r echo=TRUE}
mean_steps <- summarize(activity_by_interval, mean(steps, na.rm = TRUE))
names(mean_steps)[2] <- "mean"
```

Let's plot this variable in a time series to see how activity fluctuates throughout the day:

```{r echo=TRUE}
plot(mean_steps$interval, mean_steps$mean, type = "l", xlab = "5-Minute Interval", ylab = "Mean Number of Steps Across All Days", main = "Mean Number of Steps at 5-Minute Intervals")
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
